<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Speed comparison}
-->

# Speed comparison

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Neuroblastoma data

Consider the neuroblastoma data. There are 3418 labeled examples. If
we consider subsets, how long does it take to compute the AUM and its
directional derivatives?

```{r}
data(neuroblastomaProcessed, package="penaltyLearning")
library(data.table)
nb.err <- data.table(neuroblastomaProcessed$errors)
nb.err[, example := paste0(profile.id, ".", chromosome)]
nb.X <- neuroblastomaProcessed$feature.mat
(N.pred.vec <- as.integer(10^seq(1, 3, by=0.5)))
timing.dt.list <- list()
for(N.pred in N.pred.vec){
  N.pred.names <- rownames(nb.X)[1:N.pred]
  N.diffs.dt <- aum::aum_diffs_penalty(nb.err, N.pred.names)
  pred.dt <- data.table(example=N.pred.names, pred.log.lambda=0)
  timing.df <- microbenchmark::microbenchmark(penaltyLearning={
    roc.list <- penaltyLearning::ROChange(nb.err, pred.dt, "example")
  }, aum={
    aum.list <- aum::aum(N.diffs.dt, pred.dt$pred.log.lambda)
  }, times=10)
  timing.dt.list[[paste(N.pred)]] <- with(timing.df, data.table(
    package=expr, N.pred, seconds=time/1e9))
}
(timing.dt <- do.call(rbind, timing.dt.list))

```

Below we summarize and plot these timings.

```{r}
stats.dt <- timing.dt[, .(
  q25=quantile(seconds, 0.25),
  median=median(seconds),
  q75=quantile(seconds, 0.75)
), by=.(package, N.pred)]
library(ggplot2)
gg <- ggplot()+
  geom_line(aes(
    N.pred, median, color=package),
    data=stats.dt)+
  geom_ribbon(aes(
    N.pred, ymin=q25, ymax=q75, fill=package),
    data=stats.dt,
    alpha=0.5)+
  scale_x_log10(limits=stats.dt[, c(min(N.pred), max(N.pred)*5)])+
  scale_y_log10()
directlabels::direct.label(gg, "right.polygons")

```

From the plot above we can see that both packages have similar
asymptotic time complexity (log-linear). However aum is faster by
orders of magnitude (speedups shown below).

```{r}
stats.wide <- data.table::dcast(
  stats.dt, N.pred ~ package, value.var = "median")
stats.wide[, speedup := penaltyLearning/aum][]
```
